(Deep) Induction Rules for GADTs\vspace*{-0.1in}

### 2021-05-17 induction-paper/ENTCS-GADT-induction.tex
final MFPS

\begin{abstract} 
  Deep data types are those that are defined in terms of other data types,
  including, possibly, themselves.  In that case, they are said to be
  truly nested.  Deep induction is an extension of structural
  induction that traverses {\em all} of the structure in a deep data
  type, propagating predicates on its primitive data throughout the
  entire structure.  Deep induction was recently shown to prove
  properties of nested types, including truly nested types, that
  cannot be proved via structural induction.  In this paper we show
  how to extend deep induction to deep GADTs that are not truly
  nested. We also show that deep induction cannot be extended to truly
  nested GADTs.\looseness=-1
\end{abstract}

@@

The inductive data types handled by Coq include (possibly mutually
inductive) polynomial algebraic data types (ADTs), and the induction
rules Coq generates for them are the expected ones for standard
structural induction. It has long been understood, however, that these
rules are too weak to be genuinely useful for deep ADTs~\cite{jp20},
i.e., ADTs that are (possibly mutually inductively) defined in terms
of (other) such ADTs.\footnote{More generally, a {\em deep data type}
  is one that is (possibly mutually inductively) defined in terms of
  other such data types.}
%\footnote{Such data
%  types are called nested inductive types by Chlipala~\cite{chl},
%  reflecting the fact that ``inductive type'' means ``ADT''~in~Coq.}
Consider, for example, the following type of rose trees, here coded in
Agda and defined in terms of the standard type $\mathsf{List}$ of
lists (see Section~\ref{sec:ADTs-and-nesteds}):

@@

\[\begin{array}{l}
\mathsf{data\, Rose\, : Set \to Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;empty\, :\, Rose\,A}\\
\mathsf{\;\;\;\;\;\;\;\;\,node\,\,\,\, :\, A \to List\,(Rose\,A) \to Rose\,A} 
\end{array}\]
\noindent
The induction rule Coq automatically generates for (the analogous Coq
defintion of) rose trees is

@@

\[\begin{array}{l}
\mathsf{\forall\, (A : Set)\,(P : Rose\,A \to Set) \to P\,empty \to}\\
\mathsf{\hspace*{0.15in}
 (\forall\, (a : A)\,(ts :
  List\,(Rose\,A)) \to P\,(node\,a\,ts)) \to 
  \forall \,(x : Rose\,A) \to P\,x}
\end{array}\]
\noindent
Unfortunately, this is neither the induction rule we intuitively
expect, nor is it expressive enough to prove even basic properties of
rose trees that ought to be amenable to inductive proof. What is
needed here is an enhanced notion of induction that, when specialized
to rose trees, will propagate the predicate $\mathsf{P}$ through the
outer list structure and to the rose trees sitting inside
$\mathsf{node}$'s list argument. More generally, this enhanced notion
of induction should traverse {\em all} of the layers present in a data
structure, propagating suitable predicates to {\em all} of the data it
contains. With data types becoming ever more advanced, and with deeply
structured such types becoming increasingly ubiquitous in
formalizations, it is critically important that proof assistants
%{\color{red} (e.g., Coq, Agda, Lean)}
are able to automatically generate genuinely useful induction rules
for data types that go well beyond traditional ADTs. Such data types
include nested types~\cite{bm98}, such as $\mathsf{Rose}$ above, as
well as truly nested types\footnote{A truly nested type is a nested
  type that is defined over itself. The data type $\mathsf{Bush}$ in
  Section~\ref{sec:ADTs-and-nesteds} provides a concrete example.},
generalized algebraic data types
(GADTs)~\cite{ch03,pvww06,sp04,xcc03}, more richly indexed
families~\cite{ch88}, and deep variants of all of these.

@@

Deep induction was shown in~\cite{jp20} to be the form of induction
most appropriate to nested types (including ADTs) that are defined
over, or mutually recursively with, other such types (including,
possibly, themselves). Deep induction delivers the following genuinely
useful induction rule for rose trees:

@@

\begin{equation}\label{eq:rose}
\begin{array}{l}
\mathsf{\forall \,(A : Set)\,(P : Rose\,A \to Set)\,(Q : A \to Set)
  \to P\,empty \to}\\ 
\mathsf{\hspace*{0.15in}(\forall \,(a : A)\, (ts :
  List\,(Rose\,A))\to Q\,a \to List^\land\,P\,ts \to P\,(node\,a\,ts)) \to}\\
\mathsf{\hspace*{0.15in}\forall \,(x :
  Rose\,A) \to Rose^\land\,Q\, x \to P\,x} 
\end{array}
\end{equation}

@@

This paper shows how to extend deep induction to proper GADTs, i.e.,
to GADTs that are not simply (truly) nested types (and thus are not
ADTs).  A constructor for such a GADT $\mathsf{G}$ may, like a
constructor for a nested type, take as arguments data whose types
involve instances of $\mathsf{G}$ other than the one being defined ---
including instances that involve $\mathsf{G}$ itself. But if
$\mathsf{G}$ is a proper GADT then at least one of its constructors
will also have such a structured instance of $\mathsf{G}$ --- albeit
one not involving $\mathsf{G}$ itself --- as its codomain.
%But while the types of the arguments to a GADT's constructor can
%include instances of $\mathsf{G}$ that themselves involve
%$\mathsf{G}$, its return type cannot.
For example, the constructor $\mathsf{pair}$ for the GADT

@@

\begin{equation}\label{eq:seq}
\begin{array}{l}
\mathsf{data\, Seq\,: Set \to Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;const :\, A \to Seq\,A}\\
\mathsf{\;\;\;\;\;\;\;\;pair\;\;\, :\, Seq \,A \to Seq\,B \to
  Seq\,(A \times B)}
\end{array}
\end{equation}
\noindent
of sequences only constructs sequences of pairs, rather than sequences
of arbitrary type, as does $\mathsf{const}$.\footnote{The type of
  $\mathsf{Seq}$ is actually $\mathsf{Set \to \Set_1}$, but to aid
  readability we elide the explicit tracking of Agda universe levels in
  this paper.} If one or more of the constructors for a GADT
$\mathsf{G}$ return structured instances of $\mathsf{G}$,
%then some of
%$\mathsf{G}$'s instances might not be inhabited.
then the GADT will have two distinct, but equally natural, semantics:
a functorial semantics interpreting them as left Kan
extensions~\cite{mac71}, and a parametric semantics interpreting them
as the interpretations of their Church encodings~\cite{atk12,vw10}. As
detailed in~\cite{jgj21}, a key difference in the two semantics is
that the former views GADTs as their {\em functorial
  completions}~\cite{jp19}, and thus as containing more data than just
those expressible in syntax. By contrast, the latter views them as
what might be called {\em syntax-only} GADTs. Happily, these two views
of GADTs coincide for those that are ADTs or other nested
types. However, both they and their attendant properties differ
greatly for proper GADTs. In fact, the views deriving from the
functorial and parametric semantics for proper GADTs are sufficiently
distinct that, by contrast with the situation for ADTs and other
nested types~\cite{bfss90,gjfor15,jgj21f}, it is not actually possible
to define a functorial parametric semantics for them~\cite{jgj21}.

This observation seems, at first, to be a death knell for the prospect
of extending deep induction to GADTs. Indeed, induction can be seen as
unary parametricity, so we quickly realize that GADTs viewed as their
functorial completions cannot possibly support induction rules.  This
makes sense intuitively: induction is a syntactic proof technique, so
of course it cannot be used to prove properties of those elements of a
GADT's functorial completion that are not expressible in syntax. All
is not lost, however. As we show below, the syntax-only view of GADTs
determined by their Church encodings {\em does} support induction
rules --- including deep induction rules --- for GADTs. Perhaps
surprisingly, this paper gives the first-ever induction rules --- deep
or otherwise --- for proper GADTs. But it actually delivers far more:
it gives a general framework for deriving deep induction rules for
deep GADTs directly from their syntax. This framework can serve as a
basis for extending modern proof assistants' automatic generation of
structural induction rules for ADTs to automatic generation of deep
induction rules for GADTs. As for ADTs and other nested types, the
structural induction rule for any GADT can be recovered from its deep
induction rule simply by taking the custom predicates in its deep
induction rule to be constantly $\mathsf{True}$-valued (i.e.,
constantly $\mathsf{\top}$-valued) predicates.

Significantly, deep induction rules for GADTs cannot be derived by
somehow extending the approach of~\cite{jp20} to syntax-only
GADTs. Indeed, the approach taken there makes crucial use of the
functoriality of data types' interpretations from~\cite{jp19}, and
functoriality is precisely what interpreting GADTs as the
interpretations of their Church encodings fails to deliver. Our
approach is to instead first give a predicate lifting styled after
those of~\cite{jp20}, together with a (deep) induction rule, for the
simplest --- and arguably most important --- GADT, namely the equality
GADT. (See Section~\ref{sec:ind-equal}.)  We then derive the deep
induction rule a more complex GADT $\mathsf{G}$ by {\em i}) using the
equality GADT to represent $\mathsf{G}$ as its so-called {\em Henry
  Ford encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em ii})
using the predicate liftings for the equality GADT, and any other
GADTs appearing in the definition of $\mathsf{G}$, to appropriately
thread the custom predicates for the primitive types appearing in
$\mathsf{G}$ throughout $\mathsf{G}$'s structure. This two-step
process delivers deep induction rules for a very general class of deep
GADTs. In Section~\ref{sec:GADTs} we introduce a series of
increasingly complex GADTs as running examples, and in
Section~\ref{sec:deep-ind-GADTs} we derive a deep induction rule for
each of them. In particular, we derive the deep induction rule for
$\mathsf{Seq}$ in Section~\ref{sec:ind-seq}. We present our general
framework for deriving (deep) induction rules for (deep) GADTs in
Section~\ref{sec:framework}, and observe that the derivations in
Section~\ref{sec:deep-ind-GADTs} are all instances of it.  In
Section~\ref{sec:GADT-nested} we show that, by contrast with truly
nested types, which do have a functorial semantics, syntax-only GADTs'
lack of functoriality means that it is not possible to extend
induction --- deep or otherwise --- to truly nested GADTs. This does
not appear to be much of a restriction, however, since GADTs defined
over themselves do not, to our knowledge, appear in applications or
the literature. Section~\ref{sec:app} comprises a case study in using
deep induction. All of the deep induction rules appearing in this
paper have been derived by instantiating our general framework. Our
Agda code implementing them is available at~\cite{web-page}.

@@

{\bf Related Work\/} Various techniques for deriving induction rules
for data types that go beyond ADTs have been studied. For example, Fu
and Selinger~\cite{fs18} show, via examples, how to derive induction
rules for arbitrary nested types. Unfortunately, however, their
technique is rather {\em ad hoc}, so is unclear how to generalize it
to nested types other than the specific ones in their
examples. Moreover,~\cite{fs18} actually derives induction rules for
data types {\em related} to the original nested types rather than for
the original nested types themselves, and it is unclear whether or not
the derived rules are sufficiently expressive to prove all results
about the original nested types that we would expect to be provable by
induction. This latter point echoes the issue with Coq-derived
induction rule for rose trees raised in Section~\ref{sec:intro}, which
has the unfortunate effect of forcing users to manually write
induction (and other) rules for such types for use in that
system. Tassi~\cite{tas19} derives induction rules for data type
definitions in Coq using unary parametricity. His technique seems to
be essentially equivalent to that of~\cite{jp19} for nested types,
although he does not permit true nesting. More recently,
Ullrich~\cite{ull20} has implemented a plugin in MetaCoq to generate
induction rules for nested types. This plugin is also based on unary
parametricity and true nesting still is not permitted.  As far as we
know, no attempt has yet been made to extend either implementation to
truly nested types or to proper GADTs. In fact, we know of no work
other than that reported here that specifically addresses induction
rules for (deep) GADTs.

@@

\begin{equation*}\label{eq:list}
\begin{array}{l}
\mathsf{data\ List : Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;\;nil\,\,\,\,\,\; :\, List\,A}\\
\mathsf{\;\;\;\;\;\;\;cons\, :\, A \to List\,A \to List\,A} 
\end{array}
\end{equation*}

@@

\begin{equation*}\label{eq:ptree}
\begin{array}{l}
\mathsf{data\ PTree : Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;\;pleaf\,\,\; :\, A \to PTree\,A}\\
\mathsf{\;\;\;\;\;\;\;pnode\, :\, PTree\,(A \times A) \to PTree\,A} 
\end{array}
\end{equation*}
Perfect trees can be thought of as lists constrained to have lengths
that are powers of 2. In the above code, the constructor
$\mathsf{pnode}$ uses data of type $\mathsf{PTree\,(A \times A)}$ to
construct data of type $\mathsf{PTree\,A}$. Thus, it is clear that the
instances of $\mathsf{PTree}$ at various indices cannot be defined
independently, and that the entire inductive family of types must
therefore be defined at once. This intertwinedness of the instances of
nested types is reflected in their structural induction rules, which,
as explained in~\cite{jp20}, must necessarily involve polymorphic
predicates rather than the monomorphic predicates appearing in
structural induction rules for ADTs. The structural induction rule for
perfect trees, for example, is

@@

\[\begin{array}{l}
\mathsf{\forall (P : \forall (A : Set) \to PTree\, A \to Set)
\to \big( \forall (A : Set) (a : A) \to P\,A\,(pleaf\, a) \big) \to} \\
\quad\mathsf{\big( \forall (A : Set) (pp : PTree\,(A \times A)) \to
  P\,(A \times A)\,pp \to P\,a\,(pnode\,pp)\big) \to \forall (A :
  Set) (p : PTree\,A) \to P\,A\,p } 
\end{array}\]

@@

\[\begin{array}{l}
\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to PTree\,A
  \to Set) \to \big( \forall (A : Set) (Q : A \to Set) (a : A) \to
  Q\,a \to P\,A\,Q\,(Pleaf\, a) \big) \to} \\ \quad \mathsf{ \big(
  \forall (A : Set) (Q : A \to Set) (pp : PTree\,(A \times A)) \to
  P\,(A \times A)\,(Pair^{\wedge}\,A\,A\,Q\,Q)\,pp \to
  P\,A\,Q\,(Pnode\,pp) \big) \to} \\ \quad \mathsf{\forall (A : Set) (Q
  : A \to Set) (p : PTree\,A) \to PTree^{\wedge}\,A\,Q\,p \to
  P\,A\,Q\,p }
\end{array}\]

@@

\begin{equation*}\label{eq:bush}
\begin{array}{l}
\mathsf{data\ Bush : Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;bnil\,\,\,\,\,\; :\, Bush\,A}\\
\mathsf{\;\;\;\;\;\;bcons\, :\, A \to Bush\,(Bush\,A) \to Bush\,A} 
\end{array}
\end{equation*}

@@

\[\begin{array}{l}
\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to Bush\, A \to Set)
\to \big( \forall (A : Set) \to P\,A\,bnil \big) \to} \\ 
\quad\mathsf{\big( \forall (A : Set) (Q : A \to Set) (a : A) (bb :
  Bush\,(Bush\,A)) \to Q\,a \to
  P\,(Bush\,A)\,(Bush^{\wedge}\,A\,Q)\,bb \to P\,A\,Q\,(bcons\,a\,bb)
  \big) \to} \\ 
\quad\mathsf{\forall (A : Set) (Q : A \to Set) (b : Bush\,A) \to
  Bush^{\wedge}\,A\,Q\,b \to P\,A\,Q\,b } 
\end{array}\]

@@

Although a truly nested type admits only a single induction rule, it
is worth noting that for those nested types that do admit distinct
structural induction and deep induction rules, the latter always
generalize the former. Indeed, the structural induction rule for each
such nested type is recoverable from its deep induction rule by taking
the custom predicates on its data of primitive types to be constantly
$\mathsf{\top}$-valued predicates. This instantiation ensures that the
resulting induction rule only inspects the top-level structure of its
argument, rather than the contents of that structure, which exactly
coincides with what structural induction should do.
%A concrete example of such a derivation is given in
%Section~\ref{sec:ind-equal}.

@@

While a data constructor for a nested type can take {\em as arguments}
data whose types involve instances of that type at indices other than
the one being defined, its return type must still be at the (variable)
type instance being defined. For example, every data constructor for
$\mathsf{PTree\,A}$ returns an element of type
$\mathsf{PTree\,A}$, regardless of the instances of $\mathsf{PTree}$
appearing in the types of its arguments. GADTs relax this restriction,
allowing their data constructors both to take as arguments \emph{and
  return as results} data whose types involve instances other
than the one being defined. That is, GADTs' constructors' return type
instances can, like that of $\mathsf{pair}$ in~\eqref{eq:seq}, be
structured.\looseness=-1
%these instances can be structured.
%
%Applications of GADTs include generic programming, modeling
%programming languages via higher-order abstract syntax, maintaining
%invariants in data structures, and expressing constraints in embedded
%domain-specific languages. GADTs have also been used, e.g., to
%implement tagless interpreters~\cite{pl04,pr06,pvww06}, to improve
%memory performance~\cite{min15}, and to design APIs~\cite{pen20}.

>>>>

\begin{abstract} 
  Deep data types are those that are defined in terms of other such
  data types, including, possibly, themselves.  In that case, they are
  said to be truly nested.  Deep induction is an extension of
  structural induction that traverses {\em all} of the structure in a
  deep data type, propagating predicates on its primitive data
  throughout the entire structure.  Deep induction can be used to prove
  properties of nested types, including truly nested types, that
  cannot be proved via structural induction.  In this paper we show
  how to extend deep induction to deep GADTs that are not truly
  nested. We also show that deep induction cannot be extended to truly
  nested GADTs.\looseness=-1
\end{abstract}

@@

The inductive data types handled by Coq include (possibly mutually
inductive) polynomial algebraic data types (ADTs), and the induction
rules Coq generates for them are the expected ones for standard
structural induction. It has long been understood, however, that these
rules are too weak to be genuinely useful for deep ADTs~\cite{jp20},
i.e., ADTs that are (possibly mutually inductively) defined in terms
of (other) such ADTs.\footnote{A data type is {\em deep} if it is
  (possibly mutually inductively) defined in terms of other such data
  types (including, possibly, itself).}  The following data type of
rose trees, here coded in Agda and defined in terms of the standard
type $\mathsf{List}$ of lists (see
Section~\ref{sec:ADTs-and-nesteds}), is a deep ADT:

@@

\[\begin{array}{l}
\mathsf{data\, Rose\, : Set \to Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;empty\, :\,\forall \{A : Set\} \to Rose\,A}\\
\mathsf{\;\;\;\;\;\;\;\;\,node\,\,\,\, :\, \forall \{A : Set\} \to A \to List\,(Rose\,A) \to Rose\,A} 
\end{array}\]
\noindent
The induction rule Coq automatically generates for (the analogous Coq
defintion of) rose trees is

@@

\[\begin{array}{l}
\mathsf{\forall\, (A : Set)\,(P : Rose\,A \to Set) \to P\,empty \to}\\
\mathsf{\hspace*{0.15in}
 (\forall\, (a : A)\,(ts :
  List\,(Rose\,A)) \to P\,(node\,a\,ts)) \to 
  \forall \,(x : Rose\,A) \to P\,x}
\end{array}\]
\noindent
Unfortunately, this is neither the induction rule we intuitively
expect, nor is it expressive enough to prove even basic properties of
rose trees that ought to be amenable to inductive proof. What is
needed here is an enhanced notion of induction that, when specialized
to rose trees, will propagate the predicate $\mathsf{P}$ through the
outer list structure and to the rose trees sitting inside
$\mathsf{node}$'s list argument. More generally, this enhanced notion
of induction should traverse {\em all} of the layers present in a data
structure, propagating suitable predicates to {\em all} of the data it
contains. With data types becoming ever more advanced, and with deeply
structured such types becoming increasingly ubiquitous in
formalizations, it is critically important that proof assistants
%{\color{red} (e.g., Coq, Agda, Lean)}
are able to automatically generate genuinely useful induction rules
for data types that go well beyond traditional ADTs. Such data types
include not just deep ADTs, but also (truly\footnote{A truly nested
  type is a nested type that is defined over itself. The data type
  $\mathsf{Bush}$ in Section~\ref{sec:ADTs-and-nesteds} provides a
  concrete example.}) nested types~\cite{bm98},
%, such as $\mathsf{Rose}$ above, as well as truly nested types
generalized algebraic data types
(GADTs)~\cite{ch03,pvww06,sp04,xcc03}, more richly indexed
families~\cite{ch88}, and deep variants of all of these.

@@

Deep induction was shown in~\cite{jp20} to deliver induction rules
appropriate to deep nested types, including deep ADTs.  For the deep
ADT of rose trees, for example, it gives the following genuinely
useful induction rule:

@@

\begin{equation}\label{eq:rose}
\begin{array}{l}
\mathsf{\forall \,(A : Set)\,(P : Rose\,A \to Set)\,(Q : A \to Set)
  \to P\,empty \to}\\ 
\mathsf{\hspace*{0.15in}(\forall \,(a : A)\, (ts :
  List\,(Rose\,A))\to Q\,a \to List^\land\,(Rose\,A)\, P\,ts \to P\,(node\,a\,ts)) \to}\\
\mathsf{\hspace*{0.15in}\forall \,(x :
  Rose\,A) \to Rose^\land\,A\,Q\, x \to P\,x} 
\end{array}
\end{equation}

@@

This paper shows how to extend deep induction to proper GADTs, i.e.,
to GADTs that are not simply nested types, and thus are not ADTs.  A
constructor for such a GADT $\mathsf{G}$ may, like a constructor for a
nested type, take as arguments data whose types involve instances of
$\mathsf{G}$ other than the one being defined --- including instances
that involve $\mathsf{G}$ itself. But if $\mathsf{G}$ is a proper GADT
then at least one of its constructors will also have such a structured
instance of $\mathsf{G}$ --- albeit one not involving $\mathsf{G}$
itself --- as its codomain.
%But while the types of the arguments to a GADT's constructor can
%include instances of $\mathsf{G}$ that themselves involve
%$\mathsf{G}$, its return type cannot.
For example, the constructor $\mathsf{pair}$ for the GADT

@@

\begin{equation}\label{eq:seq}
\begin{array}{l}
\mathsf{data\, Seq\,: Set \to Set\,where}\\
\mathsf{\;\;\;\;\;\;\;\;const :\, \forall \{A : Set\} \to A \to Seq\,A}\\
\mathsf{\;\;\;\;\;\;\;\;pair\;\;\, :\,\forall \{A\,B : Set\} \to Seq \,A \to Seq\,B \to
  Seq\,(A \times B)}
\end{array}
\end{equation}
\noindent
of sequences\footnote{The type of $\mathsf{Seq}$ is actually
  $\mathsf{Set \to \Set_1}$, but to aid readability we elide the
  explicit tracking of Agda universe levels in this paper.} only
constructs sequences of pairs, rather than sequences of arbitrary
type, as does $\mathsf{const}$. If one or more of the constructors for
a GADT $\mathsf{G}$ return structured instances of $\mathsf{G}$, then
the GADT will have two distinct, but equally natural, semantics: a
functorial semantics interpreting it as a left Kan
extension~\cite{mac71}, and a parametric semantics interpreting it as
the interpretation of its Church encoding~\cite{atk12,vw10}. As
detailed in~\cite{jgj21}, a key difference in the two semantics is
that the former views GADTs as their {\em functorial
  completions}~\cite{jp19}, and thus as containing more data than just
those expressible in syntax. By contrast, the latter views them as
what might be called {\em syntax-only} GADTs. Happily, these two views
of GADTs coincide for those GADTs that are ADTs or other nested
types. However, both they and their attendant properties differ
greatly for proper GADTs. In fact, the functorial and parametric
semantics for proper GADTs are sufficiently disparate that, by
contrast with the semantics customarily given for ADTs and other
nested types~\cite{bfss90,gjfor15,jgj21f}, it is not actually possible
to define a functorial parametric semantics for them~\cite{jgj21}.

This observation seems, at first, to be a death knell for the prospect
of extending deep induction to GADTs. Indeed, induction can be seen as
unary parametricity, so we quickly realize that GADTs viewed as their
functorial completions cannot possibly support induction rules.  This
makes sense intuitively: induction is a syntactic proof technique, so
of course it cannot be used to prove properties of those elements of a
GADT's functorial completion that are not expressible in syntax. All
is not lost, however. As we show below, the syntax-only view of GADTs
determined by their Church encodings {\em does} support induction
rules --- including deep induction rules --- for GADTs.  Indeed, this
paper gives the first-ever induction rules --- deep or otherwise ---
for proper GADTs. But it actually delivers far more: it gives a
general framework for deriving deep induction rules for deep GADTs
directly from their syntax. This framework can serve as a basis for
extending modern proof assistants' automatic generation of structural
induction rules for ADTs to automatic generation of deep induction
rules for GADTs. In addition, as for ADTs and other nested types, the
structural induction rule for any GADT can be recovered from its deep
induction rule simply by taking the custom predicates in its deep
induction rule to be constantly $\mathsf{True}$-valued (i.e.,
constantly $\mathsf{\top}$-valued) predicates.

Significantly, deep induction rules for GADTs cannot be derived by
somehow extending the approach of~\cite{jp20} to syntax-only
GADTs. Indeed, the approach taken there makes crucial use of the
functoriality of data types' interpretations from~\cite{jp19}, and
functoriality is precisely what interpreting GADTs as the
interpretations of their Church encodings fails to deliver. Our
approach is to instead first give a predicate lifting styled after
those of~\cite{jp20}, together with a (deep) induction rule, for the
simplest --- and arguably most important --- GADT, namely the equality
GADT. (See Section~\ref{sec:ind-equal}.)  We then derive the deep
induction rule for a more complex GADT $\mathsf{G}$ by {\em i}) using
the equality GADT to represent $\mathsf{G}$ as its so-called {\em
  Henry Ford encoding}~\cite{ch03,hin03,mcb99,sjsv09,sp04}, and {\em
  ii}) using the predicate liftings for the equality GADT, and any
other GADTs appearing in the definition of $\mathsf{G}$, to
appropriately thread the custom predicates for the primitive types
appearing in $\mathsf{G}$ throughout $\mathsf{G}$'s structure. This
two-step process delivers deep induction rules for a very general
class of deep GADTs. In Section~\ref{sec:GADTs} we introduce a series
of increasingly complex GADTs as running examples, and in
Section~\ref{sec:deep-ind-GADTs} we derive a deep induction rule for
each of them. In particular, we derive the deep induction rule for the
$\mathsf{Seq}$ data type in~\eqref{eq:seq} in
Section~\ref{sec:ind-seq}. We present our general framework for
deriving (deep) induction rules for (deep) GADTs in
Section~\ref{sec:framework}, and observe that the derivations in
Section~\ref{sec:deep-ind-GADTs} are all instances of it.  In
Section~\ref{sec:GADT-nested} we show that, by contrast with truly
nested types, which do have a functorial semantics, syntax-only GADTs'
lack of functoriality means that it is not possible to extend
induction --- deep or otherwise --- to truly nested GADTs. This does
not appear to be much of a restriction, however, since GADTs defined
over themselves do not, to our knowledge, appear in practice or in the
literature. Section~\ref{sec:app} comprises a case study in using deep
induction. All of the deep induction rules appearing in this paper
have been derived by instantiating our general framework. Our Agda
code implementing them is available at~\cite{web-page}.

@@

{\bf Additional Related Work\/} Various techniques for deriving
induction rules for data types that go beyond ADTs have been
studied. For example, Fu and Selinger~\cite{fs18} show, via examples,
how to derive induction rules for arbitrary nested
types. Unfortunately, however, their technique is rather {\em ad hoc},
so is unclear how to generalize it to nested types other than the
specific ones studied there. Moreover,~\cite{fs18} actually derives
induction rules for data types {\em related} to the original nested
types rather than for the original nested types themselves, and it is
unclear whether or not the derived rules are sufficiently expressive
to prove all results about the original nested types that we would
expect to be provable by induction. This latter point echoes the issue
with Coq-derived induction rule for rose trees raised in
Section~\ref{sec:intro}, which has the unfortunate effect of forcing
users to manually write induction (and other) rules for such types for
use in that system. Tassi~\cite{tas19} derives induction rules for
data type definitions in Coq using unary parametricity. His technique
seems to be essentially equivalent to that of~\cite{jp19} for nested
types, although he does not permit true nesting. More recently,
Ullrich~\cite{ull20} has implemented a plugin in MetaCoq to generate
induction rules for nested types. This plugin is also based on unary
parametricity and true nesting still is not permitted.  As far as we
know, no attempts have been made to extend either implementation to
truly nested types or proper GADTs or their deep variants. In fact, we
know of no work other than that reported here that specifically
addresses induction rules for such data types.

@@

\begin{equation*}\label{eq:list}
\begin{array}{l}
\mathsf{data\ List : Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;\;nil\,\,\,\,\,\; :\, \forall \{A : Set\} \to List\,A}\\
\mathsf{\;\;\;\;\;\;\;cons\, :\, \forall \{A : Set\} \to A \to List\,A \to List\,A} 
\end{array}
\end{equation*}

@@

\begin{equation*}\label{eq:ptree}
\begin{array}{l}
\mathsf{data\ PTree : Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;\;pleaf\,\,\; :\,\forall \{A : Set\} \to A \to PTree\,A}\\
\mathsf{\;\;\;\;\;\;\;pnode\, :\,\forall \{A : Set\} \to PTree\,(A \times A) \to PTree\,A} 
\end{array}
\end{equation*}
Perfect trees can be thought of as lists constrained to have lengths
that are powers of 2. In the above code, the constructor
$\mathsf{pnode}$ uses data of type $\mathsf{PTree\,(A \times A)}$ to
construct data of type $\mathsf{PTree\,A}$. Thus, it is clear that the
instances of $\mathsf{PTree}$ at various indices cannot be defined
independently, and that the entire inductive family of types must
therefore be defined at once. This intertwinedness of the instances of
nested types is reflected in their structural induction rules, which,
as explained in~\cite{jp20}, must necessarily involve polymorphic
predicates rather than the monomorphic predicates appearing in
structural induction rules for ADTs. The structural induction rule for
perfect trees, for example, is

@@

\[\begin{array}{l}
\mathsf{\forall (P : \forall (A : Set) \to PTree\, A \to Set)
\to \big( \forall (A : Set) (a : A) \to P\,A\,(pleaf\, a) \big) \to} \\
\quad\mathsf{\big( \forall (A : Set) (pp : PTree\,(A \times A)) \to
  P\,(A \times A)\,pp \to P\,A\,(pnode\,pp)\big) \to \forall (A :
  Set) (p : PTree\,A) \to P\,A\,p } 
\end{array}\]

@@

\[\begin{array}{l}
\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to PTree\,A
  \to Set) \to \big( \forall (A : Set) (Q : A \to Set) (a : A) \to
  Q\,a \to P\,A\,Q\,(pleaf\, a) \big) \to} \\ \quad \mathsf{ \big(
  \forall (A : Set) (Q : A \to Set) (pp : PTree\,(A \times A)) \to
  P\,(A \times A)\,(Pair^{\wedge}\,A\,A\,Q\,Q)\,pp \to
  P\,A\,Q\,(pnode\,pp) \big) \to} \\ \quad \mathsf{\forall (A : Set) (Q
  : A \to Set) (p : PTree\,A) \to PTree^{\wedge}\,A\,Q\,p \to
  P\,A\,Q\,p }
\end{array}\]

@@

\begin{equation*}\label{eq:bush}
\begin{array}{l}
\mathsf{data\ Bush : Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;bnil\,\,\,\,\,\; :\, \forall \{A : Set\} \to Bush\,A}\\
\mathsf{\;\;\;\;\;\;bcons\, :\, \forall \{A : Set\} \to A \to Bush\,(Bush\,A) \to Bush\,A} 
\end{array}
\end{equation*}

@@

\[\begin{array}{l}
\mathsf{\forall (P : \forall (A : Set) \to (A \to Set) \to Bush\, A \to Set)
\to \big( \forall (A : Set)\,(Q : A \to Set) \to P\,A\,Q\,bnil \big) \to} \\ 
\quad\mathsf{\big( \forall (A : Set) (Q : A \to Set) (a : A) (bb :
  Bush\,(Bush\,A)) \to Q\,a \to
  P\,(Bush\,A)\,(P\,A\,Q)\,bb \to P\,A\,Q\,(bcons\,a\,bb)
  \big) \to} \\ 
\quad\mathsf{\forall (A : Set) (Q : A \to Set) (b : Bush\,A) \to
  Bush^{\wedge}\,A\,Q\,b \to P\,A\,Q\,b } 
\end{array}\]

@@

Although a truly nested type admits only a single induction rule, it
is worth noting that for those nested types that do admit distinct
structural induction and deep induction rules, the latter always
generalize the former. Indeed, the structural induction rule for each
such nested type is recoverable from its deep induction rule by taking
the custom predicates on its data of primitive types to be constantly
$\mathsf{\top}$-valued predicates. This instantiation ensures that the
resulting induction rule only inspects the top-level structure of its
argument, rather than the contents of that structure, which is exactly
what structural induction should do.
%A concrete example of such a derivation is given in
%Section~\ref{sec:ind-equal}.

@@

While a data constructor for a nested type can take {\em as arguments}
data whose types involve instances of that type at indices other than
the one being defined, its return type must still be at the (variable)
type instance being defined. For example, each of $\mathsf{pleaf}$ and
$\mathsf{pnode}$ returns an element of type $\mathsf{PTree\,A}$
regardless of the instances of $\mathsf{PTree}$ appearing in the types
of its arguments. GADTs relax this restriction, allowing their data
constructors both to take as arguments \emph{and return as results}
data whose types involve instances other than the one being
defined. That is, GADTs' constructors' return type instances can, like
that of $\mathsf{pair}$ in~\eqref{eq:seq}, be structured.\looseness=-1
%these instances can be structured.
%
%Applications of GADTs include generic programming, modeling
%programming languages via higher-order abstract syntax, maintaining
%invariants in data structures, and expressing constraints in embedded
%domain-specific languages. GADTs have also been used, e.g., to
%implement tagless interpreters~\cite{pl04,pr06,pvww06}, to improve
%memory performance~\cite{min15}, and to design APIs~\cite{pen20}.

GADTs are used in precisely those situations in which different
behaviors at different instances of data types are desired. This is
achieved by allowing the programmer to give the type signatures of the
GADT's data constructors independently, and then taking advantage of
pattern matching to force the desired type refinement. For example,
the {\em equality} GADT
\begin{equation}\label{eq:equal}
\begin{array}{l}
\mathsf{data\ Equal : Set \to Set \to Set\ where}\\
\mathsf{\;\;\;\;\;\;\;\;\;\;\;\,refl :\, \forall \{A : Set\} \to Equal\,A\,A}
\end{array}
\end{equation}
is parameterized by two type indices, but it is only possible to
construct data elements of type $\mathsf{Equal\,A\,B}$ if $\mathsf{A}$
and $\mathsf{B}$ are instantiated at the same type. If the types
$\mathsf{A}$ and $\mathsf{B}$ are syntactically identical then the
type $\mathsf{Equal\,A\,B}$ contains the single data element
$\mathsf{refl}$. It contains no data elements otherwise.
